---
title: "SlidyTest"
author: "Jeff Newmiller"
date: "September 29, 2018"
output: slidy_presentation
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
library(DiagrammeR)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Notes

https://stackoverflow.com/questions/20847371/two-column-layouts-in-rstudio-presentations-slidify-pandoc

## Test

<div style="float: left; width: 50%;">

https://community.rstudio.com/t/r-presentations-for-beginners/1222/3

 - some
 - text 
 - here

</div>
<div style="float: right; width: 50%;">

- Column two

</div>

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r,tidy=TRUE}
dta <- read.csv( "../data/MAC000002clean.csv"
               , as.is=TRUE # avoid auto-convert to factor
               )
dta$Dtm <- as.POSIXct( dta$DateTime )
str( dta )
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## Good Practices

Converting character to `POSIXct`:

```{r,echo=FALSE,fig.height=3}
mermaid('
sequenceDiagram
  File ->> Character : read.csv(as.is=TRUE)
  Character ->> NA : as.POSIXct("2013-21-03 02:30:00")
  Character ->> POSIXct : as.POSIXct("2013-03-10 14:30:00")
  Character ->> POSIXct : as.POSIXct("3/21/13 2:30pm",format="%m/%d%y %I:%M%p")
')
```

## Trouble with `POSIXct`

```{r,echo=FALSE,fig.height=3}
mermaid('
sequenceDiagram
  File ->> Factor : read.csv(as.is=FALSE)
  Factor ->> InvalidPOSIXct : as.POSIXct(fctr)
  Factor ->> Character : as.character(fctr)
')
```

* In `read.csv`, the `stringsAsFactors` parameter gets set to `TRUE` when `as.is=FALSE`.
* 

## Windows Spring Forward Invalid Time

```{r,fig.height=3,fig.width=7}
Sys.setenv( TZ = "US/Pacific" )
dta1 <- read.csv( "../data/RawSpringForward.csv", as.is=TRUE )
dta1$Dtm <- as.POSIXct( dta1$Timestamp, format="%Y-%m-%d %H:%M:%S" )
dta1$Index <- seq.int( nrow( dta1 ) )
plot( Dtm ~ Index, data=dta1, type="l" )
```

## Windows Spring Forward Invalid Time

```{r}
dta1
```

