---
title: "More Datetime Howto"
author: "Jeff Newmiller"
date: "September 9, 2018"
output: slidy_presentation
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
```

## Other approaches to handling time

Some people think the `POSIXt` approach is too rigid, and try out their own ideas for handling time:

- `chron::chron`
    * Contributed package, behaves similar to Excel
    * Internally represented as days since origin date (1970-01-01), a floating point number, including the fractional part!
- `zoo::yrmon`
    * Cannot handle increments of less than one month
    * Internally represented as numeric years, a floating point number, always rounded to nearest 1/12 when displayed

## Detour: Floating Point Error

R FAQ 7.31 warns against depending on exact results when using floating point fractions:

```{r}
x <- 0.3
0.6 == 2 * x # works
0.9 == 3 * x # but you cannot rely on it
0.9 - 3 * x
```

This imprecision is not unique to R... this applies to all software that uses floating point numbers.

It is best to use a date/time representation that uses non-fractional values for your application... `Date` if you never use time-of-day, or `POSIXct` if your smallest precision is one second.  If your smallest precision is less than one second, `POSIXt` may introduce rounding errors so it is best to minimize the amount of calculations performed with such timestamps.

## `chron`

- Very similar to Excel time (simple)

```{r,echo=TRUE}
library(chron)
dtm1 <- chron( dates. = c( "3/13/2016", "3/13/2016" )
             , times. = c( "01:00:00", "03:00:00" )
             )
dtm1
as.numeric( dtm1 )
dtm1[ 1 ] < dtm1[ 2 ]
diff( dtm1 )
```

## `chron` Spring Forward

- Does not understand timezones
- base graphics plotting doesn't display right (ggplot does)

```{r,fig.height=3}
library(ggplot2)
dtms1 <- seq( dtm1[ 1 ], dtm1[ 2 ], times( "00:30:00" ) ); dtms1
qplot( seq_along( dtms1 ), dtms1 ) + chron::scale_y_chron(format="%m/%d/%y %H:%M" )
```

## `chron` Sequence Rounding

```{r}
dtm2a <- chron( "02/20/13", "00:00:00" )
dtm2b <- chron( "07/03/18", "15:30:00" )
dtm2 <- seq( from=dtm2a, to=dtm2b, by=times( "00:15:00" ) )
tail( dtm2 )
length( dtm2 )
Sys.setenv( TZ="US/Pacific" )
dtm3a <- as.POSIXct( "02/20/13 00:00:00", format = "%m/%d/%y %H:%M:%S" )
dtm3b <- as.POSIXct( "07/03/18 15:30:00", format = "%m/%d/%y %H:%M:%S" )
dtm3 <- seq( from = dtm3a, to=dtm3b, by=as.difftime( 15, units="mins" ) )
tail( dtm3 )
length( dtm3 )
```

